\Large\textbf{Preparation GUIDE for Initial-Submission Manuscript}\\~\\ \large\textbf{ Preparation of Papers in Two-Column Format\\ for the ASP-DAC 2014 (\LaTeX2e version)}

### 2013-12-04 delays.tex
Updated the paper

The \texttt{delay(M..M)} variant is a little more interesting. Like
before we find sets $S_1$ and $S_2$, and find the intersection of the
two sets to get the value of \texttt{d}. It is possible (and often
likely, as suggested by our experiments in
Section~\ref{sec:experimental-results}) that the resultant set $D$ is
empty (also possible in the case of \texttt{delay(M..N)}, but never
possible in case of \texttt{delay(M)}). Note that $S_1$ and $S_2$ can
never be empty sets and will always contain at minimum a single
element. In such a case, we automatically \textit{relax} the upper bound
of the \texttt{delay} statement.

>>>>

The \texttt{delay(M..M)} variant is more interesting. Like before, we
find sets $S_1$ and $S_2$, and find the intersection of the two sets to
get the value of \texttt{d}. It is possible (and often likely, as
suggested by our experiments in Section~\ref{sec:experimental-results})
that the resultant set $D$ is empty (also possible in the case of
\texttt{delay(M..N)}, but never possible in case of
\texttt{delay(M)}). Note that $S_1$ and $S_2$ can never be empty sets
and will always contain at minimum a single element. In such a case, we
can \textit{relax} the upper bound of the \texttt{delay} statement.

### 2013-12-04 discussion.tex
Updated exps
Updated the paper

Many real-world systems require non-deterministic timing
constructs. Where the exact real-time delay is not known or should not
be known apriori. One such system is our motivating example -- the human
response time system (HRCTS). Another one we presented was a real
printer-spooler embedded controller in
Section~\ref{sec:experimental-results}.

@@

There are many instances when one would like to wait on an input from
environment for only a specified amount of time. This can be programmed
as:
% \begin{scriptsize}
\begin{verbatim}
// timeout after 1ms.
trap(T){{await(A);}||
        {delay(1 ms);exit(T);}}
\end{verbatim}
% \end{scriptsize}

In the above case, the program waits for signal \texttt{A} from the
environment for \textit{at\_least} \texttt{1 ms}, but it may wait longer
if other reactions/CDs are scheduled for execution. This second form of
delay statement is very useful for programming low priority tasks.

@@

The third form, is useful for hard-real time guarantees, e.g., a
reaction, or a whole CD, can be programmed to run periodically like so:

\begin{verbatim}
while(true) { delay ((1..1)ms); emit S; 
 //do something }
\end{verbatim}

@@

\begin{verbatim}
while(true) {      while(true) {
 delay(1..1 ms);    while(true) { 
 //extra pause
 pause;              trap(T) {
 emit S;              int x = 0;
}                     while(true) {
                       x = x + 1;
                       pause;
                       if (x == d)exit(T);
                      }
                    }
                    //extra pause
                    pause;
                    emit S;
                   }
\end{verbatim}

The system designer needs to be aware that just introducing a delay
construct does not make the reaction (or CD) periodic. The program on
the left does \textit{not} emit signal \texttt{S} every \texttt{1 ms},
because of the extra \texttt{pause} statement. Introducing additional
pauses introduces more logical ticks, consequently increasing the
real-time delay.

@@

% \begin{scriptsize}
\begin{verbatim}
suspend(S) {emit START_TIMER(10); 
            await (TIMER);emit O1;}
\end{verbatim}
% \end{scriptsize}

@@

\subsection{Other semantic discussion}
\label{sec:other-semant-dissc}

\subsubsection{Interaction of channel communication and delays}
\label{sec:inter-chann-comm}

>>>>

Many real-world systems require non-deterministic timing
constructs. Where the exact real-time delay is not known or should not
be known apriori. One such system is our motivating example -- the human
response time system (HRTCS). Another one we presented was a real
printer-spooler embedded controller in
Section~\ref{sec:experimental-results}.

@@

There are many instances when one would like to wait on an input from
environment for only a specified amount of time. This can be programmed
as:

\begin{figure}[h!]
  \centering
    \begin{minipage}[b]{\linewidth}
\begin{verbatim}
// timeout after 1ms.
trap(T){{await(A);}||
        {delay(1 ms);exit(T);}}
\end{verbatim}
    \end{minipage}
  \caption{Programming low priority timeout tasks with delay statements}
  \label{fig:timeout}
\end{figure}

In the above case, the program waits for signal \texttt{A} from the
environment for \textit{at least} \texttt{1 ms}, but it may wait longer
if other reactions/CDs are scheduled for execution. This second variant
of delay statement is very useful for programming low priority tasks.

@@

The third variant is useful for hard-real time guarantees, e.g., a
reaction, or a whole CD, can be programmed to run periodically like so:

\begin{figure}[h!]
  \centering
  \begin{minipage}[b]{\linewidth}
\begin{verbatim}
while(true) { delay ((1..1)ms); emit S; 
 //do something }
\end{verbatim}
  \end{minipage}
  \caption{Programming periodic tasks with delay statements}
  \label{fig:periodic}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{SubFloat}{\label{pp:a}Original program}
    \centering
    \begin{minipage}[b]{0.3\linewidth}
\begin{verbatim}
while(true) {
 delay(1..1 ms);
 //extra pause
 pause;
 emit S;
}
\end{verbatim}
  \end{minipage}
\end{SubFloat}
\hspace{9pt}
  \begin{SubFloat}{\label{pp:b}Rewritten program}
    \begin{minipage}[h]{0.5\linewidth}
\begin{verbatim}
while(true) {
 while(true) { 
  trap(T) {
   int x = 0;
   while(true) {
    x = x + 1;
    pause;
    if (x == d)
     exit(T);
   }
  }
  //extra pause
  pause;
  emit S;
 }
}
\end{verbatim}
    \end{minipage}
  \end{SubFloat}
\caption{Extra pause statements in periodic tasks}
  \label{fig:periodic2}
\end{figure}

The system designer needs to be aware that just introducing a delay
construct does not make the reaction (or CD) periodic. The program on
the left does \textit{not} emit signal \texttt{S} every \texttt{1 ms},
because of the extra \texttt{pause} statement. Introducing additional
pauses introduces more logical ticks, consequently increasing the
real-time delay. Static analysis techniques from real-time community can
be used to automatically determine the delay specification in such
cases, but describing these techniques is out of the scope of this work.

@@

\begin{figure}[h!]
  \centering
  \begin{minipage}[b]{\linewidth}
\begin{verbatim}
suspend(S) {
 emit START_TIMER(10); 
 await (TIMER);emit O1;
}
\end{verbatim}
  \end{minipage}
  \caption{Interaction of preemption and external timers}
  \label{fig:preemp}
\end{figure}

@@

% \subsection{Other semantic discussion}
% \label{sec:other-semant-dissc}

\subsection{Interaction of channel communication and delays}
\label{sec:inter-chann-comm}

### 2013-12-04 exps.tex
Updated the paper

Motor is a stepper motor controller from~\cite{Bourke2009a}, used for
producing monochrome images on paper by applying a correct application
of current to a print head (a row of resistors). In this example,
duration of the current applied is controlled by the feedback logic
using real-time delays, which are expressed using reactive
constructs. Printer is an example from~\cite{Schneider:1999:CRT:555233},
which models a printer and a spooler in timed CSP.  Their operation is
rather simple: as soon as the spooler receives a print job, passes it to
the printer through a channel where actual printing process take
place. This example along with HRCTS demonstrates the use of
non-deterministic delays.

@@

HRTCS consists of 2 CDs, but only one with real-time delays. AECS also
has 2 CDs; CD1 consist of a single delay statement, while CD2 has 2
delay statements. In this case, as explained previously, delay
statements replace the external timers and corresponding interface (I/O)
signals. The system triggers an alarm 10 seconds after detection of an
intruder or generates a beep sound when an entrance door is being opened
for the same amount of time. In the Printer example, spooler and printer
processes are mapped to separate CDs. Spooler CD sends a job to the
Printer CD between 2.7 to 10.6 ms after receiving it from the
environment (e.g. a user). On the other hand, Printer CD notifies
environment that it starts printing between 1.2 to 30.3 ms once it
receives the job from the Spooler CD. Actual print job is started
between 5.6 to 100.2 ms after the notification. In this example, bounded
delay ensures that both channel communication and print job should occur
within the specified amount of time. Lastly, Motor is a synchronous
program with 5 delay statements that control the amount of current
applied to the stepper motor coils.

@@

We use the \textit{Java Optimized Processor}
(JOP)~\cite{jop:jnl:jsa2007}, to carry out all our experiments. JOP
implements the JVM specification in hardware directly, and has been
shown to be real-time amenable. We statically estimate the WCRT and BCRT
of these benchmark applications using the real-time analysis tools
provided by the JOP design framework. The \textit{Safety Critical Java}
(SCJ) specification we use for real-time analysis avoids garbage
collection (GC) and hence, GC overheads and unpredictability are
completely avoided.

@@

The previous results show that HRTCS and Printer were the only examples
that the compiler was able to find \textit{d} without relaxing upper
bound of \texttt{delay}. Figure~\ref{fig:graph1} shows the accuracy of
our results by comparing differences between programmer specified upper
bounds (Table~\ref{fig:benchlist}) and the actual bounds obtained using
our techniques (Table~\ref{fig:comparison}).

\begin{figure}[h!]
	\centering
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x
			coords={HRTCS,Printer/CD1,Printer/CD2,AECS/CD2},
			height=5cm,
			width=9.5cm,
			minor y tick num=1,
			yminorgrids=true,
			ymajorgrids=true,
			ytick={0,0.5,...,2},
			enlarge x limits=0.3,
%      xlabel=Benchmark,
			x tick label style={rotate=45,font=\footnotesize},
			ylabel=Difference factor,
		]
			\addplot coordinates
			{(HRTCS,1)(Printer/CD1,1)(Printer/CD2,1)(AECS/CD2,1.94)};
			\addplot coordinates {(Printer/CD2, 1)(AECS/CD2, 1.94)};
	\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x coords={Motor},
			ytick={0,0.5,...,5},
			ymin={1.5},
			minor y tick num=1,
			yminorgrids=true,
			ymajorgrids=true,
%      extra y ticks={2.5},
%      extra y tick labels={asdf},
%      extra y tick style={grid=major},
			height=5cm,
			width=9.5cm,
			xlabel=Benchmark,
			xlabel shift={4pt},
			ylabel=Difference factor,
			legend style={
				anchor=north,
				at={(0.45,-0.44)},
				legend columns=-1,
			}
%      nodes near coords
		]
			\addplot coordinates {(Motor, 2.60)};
			\addplot coordinates {(Motor, 2.63)};
			\addplot coordinates {(Motor, 4.39)};
			\addplot coordinates {(Motor, 2.92)};
			\addplot coordinates {(Motor, 2.69)};
			\legend{Delay1,Delay2,Delay3,Delay4,Delay5}
	\end{axis}
\end{tikzpicture}
\caption{Accuracy of the results}
\label{fig:graph1}
\end{figure}

@@

In Figure~\ref{fig:improv-a}, the program simply checks for the id of a
person entering or exiting an intelligent room through the signals
called \texttt{EntryRequest} and \texttt{ExitRequest}, and verifies the
id using data logic written in Java (\texttt{isValidRequest}). We
identified this execution path caused lengthening the WCRT of the
program. In this case, we divide two consequtive statements, i.e.
\texttt{present} and \texttt{if}, by inserting \texttt{pause} such that
each of these statement is executed in two different ticks. The same
technique is applied to Motor example, which is purely
control-dominated, to reduce WCRT of the program. Note that we can also
optimize algorithms of the data computations used in each Java method to
reduce WCRT. However it is out of scope of this paper, hence not
discussed here.

The result of the upper bound relaxation with reduced WCRT is shown in
Figure~\ref{fig:graph2}. As one can see from the Figure, the upper
bounds for delay statements in AECS/CD2 have been decreased by 11.81\%
which is higher than all those in Motor which ranges from 7.54 to
8.05\%. It was as expected since amount of computation required
in control flow is mostly less demanding than data such that breaking
the critical paths in Motor did not reduce WCRT as much as AECS/CD2.
Therefore, we can see from the result that the size of the relaxation
depends on how designers write their program as well as typs of the
application (i.e. data or control dominated).

@@

\begin{figure}[ht!]
	\centering
%\begin{tikzpicture}[scale=0.9]
%  \begin{axis}[
%      ybar,xtick=data,
%      symbolic x
%      coords={AECS/CD2},
%      height=5cm,
%      width=9.5cm,
%      minor y tick num=1,
%      yminorgrids=true,
%      ymajorgrids=true,
%      xlabel=Benchmark,
%      ylabel=Percentage reduction,
%    ]
%      \addplot coordinates
%      {(AECS/CD2, 11.82)};
%  \end{axis}
%\end{tikzpicture}
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x coords={Motor,AECS/CD2},
			ytick={0,1,...,12},
			ymin={6},
			minor y tick num=1,
			enlarge x limits=0.5,
			enlarge y limits=0.1,
			yminorgrids=true,
			ymajorgrids=true,
			height=7cm,
			width=9.5cm,
			xlabel=Benchmark,
			xlabel shift={4pt},
			ylabel=Percentage reduction (\%),
			legend style={
				anchor=north,
				at={(0.45,-0.29)},
				legend columns=-1,
			}
		]
			\addplot coordinates {(Motor, 8.05) (AECS/CD2, 11.82)};
			\addplot coordinates {(Motor, 8.00)};
			\addplot coordinates {(Motor, 6.42)};
			\addplot coordinates {(Motor, 7.53)};
			\addplot coordinates {(Motor, 7.89)};
			\legend{Delay1,Delay2,Delay3,Delay4,Delay5}
	\end{axis}
\end{tikzpicture}
\caption{Reduction in upper bound relaxation with improved WCRT}
\label{fig:graph2}
\end{figure}

>>>>

Motor is a stepper motor controller from~\cite{Bourke2009a}, used for
producing monochrome images on paper by correct application of current
to a print head (a row of resistors). In this example, duration of the
current applied is controlled by the feedback logic using real-time
delays, which are expressed using reactive constructs. Printer is an
example from~\cite{Schneider:1999:CRT:555233}, which models a printer
and a spooler in timed CSP. Its operation can be described as follows:
as soon as the spooler receives a print job it passes it to the printer
through a channel where actual printing process take place. This example
along with HRTCS demonstrates the use of non-deterministic delays.

@@

HRTCS consists of 2 CDs, but only one with real-time delays. AECS also
has 2 CDs; CD1 consist of a single delay statement, while CD2 has 2
delay statements. In this case, as explained previously, delay
statements replace the external timers and corresponding interface (I/O)
signals. The system triggers an alarm 10 seconds after detection of an
intruder or generates a beep sound when an entrance door is being opened
for the same amount of time. In the Printer example, spooler and printer
processes are mapped to separate CDs. Spooler CD sends a job to the
Printer CD between 2.7 to 10.6 ms after receiving it from the
environment (e.g. a user). On the other hand, Printer CD notifies the
environment that it starts printing between 1.2 to 30.3 ms once it
receives the job from the Spooler CD. Actual print job is started
between 5.6 to 100.2 ms after the notification. In this example, bounded
delay ensures that both; channel communication and print job should
occur within the specified amount of time. Lastly, Motor is a
synchronous program with 5 delay statements that control the amount of
current applied to the stepper motor coils.

@@

We use the \textit{Java Optimized Processor}
(JOP)~\cite{jop:jnl:jsa2007} to carry out all our experiments. JOP
implements the JVM specification in hardware directly and has been shown
to be real-time amenable. We statically estimate the WCRT and BCRT of
these benchmark applications using the real-time analysis tools provided
by the JOP design framework. The \textit{Safety Critical Java} (SCJ)
specification we use for real-time analysis avoids garbage collection
(GC) and hence, GC overheads and unpredictability are completely
avoided.

@@

The previous results show that for HRTCS and Printer that the compiler
was able to find \textit{d} without relaxing upper bound of
\texttt{delay}. Figure~\ref{fig:graph1} shows the accuracy of our
results by comparing differences between programmer specified upper
bounds (Table~\ref{fig:benchlist}) and the actual bounds obtained using
our techniques (Table~\ref{fig:comparison}) for the other examples
requiring relaxation. AECS/CD1 delay is not included in
Figure~\ref{fig:graph2}, because although it requires relaxation, the
delay specification (Table~\ref{fig:benchlist}) does not have an upper
bound.

\begin{figure}[h!]
	\centering
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x
			coords={HRTCS,Printer/CD1,Printer/CD2,AECS/CD2},
			height=5cm,
			width=9.5cm,
			minor y tick num=1,
			yminorgrids=true,
			ymajorgrids=true,
			ytick={0,0.5,...,2},
			enlarge x limits=0.3,
%      xlabel=Benchmark,
			x tick label style={rotate=45,font=\footnotesize},
			ylabel=Difference factor,
		]
			\addplot + [postaction={pattern=north east lines}] coordinates
			{(HRTCS,1)(Printer/CD1,1)(Printer/CD2,1)(AECS/CD2,1.94)};
			\addplot + [postaction={pattern=north west
                          lines}] coordinates 
                        {(Printer/CD2, 1)(AECS/CD2, 1.94)};
	\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x coords={Motor},
			ytick={0,0.5,...,5},
			ymin={1.5},
			minor y tick num=1,
			yminorgrids=true,
			ymajorgrids=true,
%      extra y ticks={2.5},
%      extra y tick labels={asdf},
%      extra y tick style={grid=major},
			height=5cm,
			width=9.5cm,
			xlabel=Benchmark,
			xlabel shift={4pt},
			ylabel=Difference factor,
			legend style={
				anchor=north,
				at={(0.45,-0.44)},
				legend columns=-1,
			}
%      nodes near coords
		]
			\addplot + [postaction={pattern=north east lines}] coordinates
                        {(Motor, 2.60)};
			\addplot + [postaction={pattern=north west lines}] coordinates
                        {(Motor, 2.63)};
			\addplot + [postaction={pattern=dots}] coordinates
                        {(Motor, 4.39)};
			\addplot + [postaction={pattern=horizontal lines}] coordinates
                        {(Motor, 2.92)};
			\addplot + [postaction={pattern=crosshatch}] coordinates
                        {(Motor, 2.69)};
			\legend{Delay1,Delay2,Delay3,Delay4,Delay5}
	\end{axis}
\end{tikzpicture}
\caption{Accuracy of the results}
\label{fig:graph1}
\end{figure}

@@

In Figure~\ref{fig:improv-a}, the program simply checks for the id of a
person entering or exiting an intelligent room through the signals
called \texttt{EntryRequest} and \texttt{ExitRequest}, and verifies the
id using data logic written in Java (\texttt{isValidRequest}). We
identified this as the critical execution path accounting for the WCRT.
In this case, we divide two consequtive statements, i.e.
\texttt{present} and \texttt{if}, by inserting \texttt{pause} such that
each of these statement is executed in two different ticks. The same
technique is applied to Motor example, which is purely
control-dominated, to reduce WCRT of the program. Note that we can also
optimize algorithms of the data computations used in each Java method to
reduce WCRT. However it is out of scope of this paper, hence not
discussed here.

The result of the upper bound relaxation with reduced WCRT is shown in
Figure~\ref{fig:graph2}.  The upper bounds for delay statements in
AECS/CD2 decreased by 11.81\%, while that for Motor example reduced from
7.54 to 8.05\%. This is expected since the amount of computation
required in control flow is less demanding than data and hence, breaking
the critical paths in Motor did not reduce WCRT as much as AECS/CD2.
Therefore, we can see from the result that the size of the relaxation
depends on how designers write their program as well as type of the
application (i.e. data or control dominated).

@@

\begin{figure}[ht!]
	\centering
%\begin{tikzpicture}[scale=0.9]
%  \begin{axis}[
%      ybar,xtick=data,
%      symbolic x
%      coords={AECS/CD2},
%      height=5cm,
%      width=9.5cm,
%      minor y tick num=1,
%      yminorgrids=true,
%      ymajorgrids=true,
%      xlabel=Benchmark,
%      ylabel=Percentage reduction,
%    ]
%      \addplot coordinates
%      {(AECS/CD2, 11.82)};
%  \end{axis}
%\end{tikzpicture}
\begin{tikzpicture}[scale=0.9]
	\begin{axis}[
			ybar,xtick=data,
			symbolic x coords={Motor,AECS/CD2},
			ytick={0,1,...,12},
			ymin={6},
			minor y tick num=1,
			enlarge x limits=0.5,
			enlarge y limits=0.1,
			yminorgrids=true,
			ymajorgrids=true,
			height=7cm,
			width=9.5cm,
			xlabel=Benchmark,
			xlabel shift={4pt},
			ylabel=Percentage reduction (\%),
			legend style={
				anchor=north,
				at={(0.45,-0.29)},
				legend columns=-1,
			}
		]
			\addplot + [postaction={pattern=north east lines}] coordinates
                        {(Motor, 8.05) (AECS/CD2, 11.82)};
			\addplot + [postaction={pattern=north west lines}] coordinates
                        {(Motor, 8.00)};
			\addplot + [postaction={pattern=dots}] coordinates
                        {(Motor, 6.42)};
			\addplot + [postaction={pattern=horizontal lines}] coordinates
                        {(Motor, 7.53)};
			\addplot + [postaction={pattern=crosshatch}] coordinates
                        {(Motor, 7.89)};
			\legend{Delay1,Delay2,Delay3,Delay4,Delay5}
	\end{axis}
\end{tikzpicture}
\caption{Reduction in upper bound relaxation with improved WCRT}
\label{fig:graph2}
\end{figure}

### 2013-12-04 motivation.tex
Updated the paper

Consider a simple machine used to test human response time and collect
this data for future research purposes. The machine switches on a green
light on a touch screen for anywhere between 50.3 ms to 200.3 ms, if the
user can touch this green light then she is successful and an integer
\texttt{1} is sent to a database, if unsuccessful a \texttt{0} is sent
instead, Different statistics and analyses on human responsiveness can
be based on collected data.

@@

The SystemJ program is divided into two clock-domains, the first
clock-domain continuously displays a green light (\texttt{GREEN\_LIGHT}
signal) on the touch screen sensor and waits for the user to respond. If
the user is able to touch the screen within the specified time of 50.3
ms to 200.3 ms a positive response is sent to the second
clock-domain. All the SystemJ programming constructs used to implement
this system are described in~\cite{amal10} and Table~\ref{tab:1} in
Section~\ref{sec:background}. The delay speciﬁcation, which models the
non-deterministic delay between 50.3 ms and 200.3 ms is not supported in
the current language and cannot be compiled with the current SystemJ
compiler. We introduce real-time \texttt{delay} statements in SystemJ.

>>>>

Consider a machine used to test human response time and collect this
data for future research purposes. The machine switches on a green light
on a touch screen for anywhere between 50.3 ms to 200.3 ms, if the user
can touch this green light then she is successful and an integer
\texttt{1} is sent to a database, if unsuccessful a \texttt{0} is sent
instead. Different statistics and analyses on human responsiveness can
be based on the collected data.

@@

The SystemJ program is divided into two clock-domains, the first
clock-domain continuously displays a green light (\texttt{GREEN\_LIGHT}
signal) on the touch screen sensor and waits for the user to respond. If
the user is able to touch the screen within the specified time of 50.3
ms to 200.3 ms a positive response is sent to the second
clock-domain. All the SystemJ programming constructs used to implement
this system are described in~\cite{amal10} and Table~\ref{tab:1} in
Section~\ref{sec:background}. The delay specification, which models the
non-deterministic delay between 50.3 ms and 200.3 ms is not supported in
the current language and cannot be compiled with the current SystemJ
compiler. We introduce real-time \texttt{delay} statements in SystemJ.
